---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bayesVC

<!-- badges: start -->
<!-- badges: end -->

bayesVC implements two nonparametric variable selection methods for Bayesian tree ensembles:

* **DART VC-measure** — introduced in [*Posterior Summarization for Variable Selection in Bayesian Tree Ensembles*](https://arxiv.org/abs/2509.07121). Extracts the Variable Count (VC) from DART fits and applies Hierarchical Agglomerative Clustering (HAC) to separate signal variables from noise.

* **DART VIP Rank** — introduced in [Ab Initio Nonparametric Variable Selection for Scalable Symbolic Regression with Large $p$](https://openreview.net/forum?id=9gyJJw8ZUj). Extracts the Variable Inclusion Proportion (VIP) from DART fits, ranks variables by average VIP, and applies HAC as a pre-screening step to identify candidate predictors with high sensitivity.

Both methods support **regression** and **binary classification** (experimental) tasks, and integrate with any DART backend via the [`future`](https://future.futureverse.org/) package for parallel execution.

## Installation

You can install the development version of `bayesVC` from [GitHub](https://github.com/mattsheng/bayesVC) with:

``` r
# install.packages("pak")
pak::pkg_install("mattsheng/bayesVC")
```

## Backends

The bayesVC R package currently supports two flavors of DART implementations:

* `BART`: installed automatically with `bayesVC`, or can be easily installed via `install.packages('BART')`
* `dartMachine`: requires installation from [https://github.com/theodds/dartMachine](https://github.com/theodds/dartMachine)

Note: the `dartMachine` backend is only supported for regression (`mode = "regression"`).

## Example: DART VC-measure

We demonstrate DART VC-measure using the Friedman equation:

$$y = 10\sin(\pi x_1x_2) + 20(x_3-0.5)^2 + 10x_4 + 5x_5 + \varepsilon,$$

where $\varepsilon \sim N(0, 2^2)$. Predictors $x_1,\ldots,x_{100}$ are sampled iid from an Uniform(0, 1) distribution. The goal is to identify the truly relevant predictors, namely $(x_1,x_2,x_3,x_4,x_5)$, from all $p=100$ predictors.

### Regression

```{r example, message = FALSE}
library(bayesVC)

set.seed(123)
n <- 1000
p <- 100
Lrep <- 10

# Generate data (y, X)
X <- matrix(runif(n * p), n, p)
y_mu <- 10 * sin(pi * X[, 1] * X[, 2]) + 20 * (X[, 3] - 0.5)^2 + 10 * X[, 4] + 5 * X[, 5]
eps <- rnorm(n, mean = 0, sd = 2)
y <- y_mu + eps

# Perform DART VC-measure using `BART` backend
VC_result <- DartVC(
  y = y,
  X = X,
  seed = 123,
  Lrep = Lrep,
  backend = "BART",
  mode = "regression"
)

# Examine the selected predictors
VC_result$pos_idx
```

### Binary classification

Pass `mode = "classification"` to switch to probit BART (`BART::pbart()`). The response `y` must be a numeric vector with exactly 2 unique values.

```{r classification, message = FALSE}
library(bayesVC)

set.seed(123)
n <- 500
p <- 100
Lrep <- 10

# Generate binary response via Friedman probit DGP
X <- matrix(runif(n * p), n, p)
mu <- 10 * sin(pi * X[, 1] * X[, 2]) + 20 * (X[, 3] - 0.5)^2 + 10 * X[, 4] + 5 * X[, 5]
mu_scaled <- (mu - mean(mu)) / sd(mu)
y <- rbinom(n, size = 1, prob = pnorm(mu_scaled))

# Perform DART VC-measure for binary classification
VC_result <- DartVC(
  y = y,
  X = X,
  seed = 123,
  Lrep = Lrep,
  backend = "BART",
  mode = "classification"
)

VC_result$pos_idx
```

## Example: DART VIP Rank

DART VIP Rank (`DartVIP`) is designed as a **pre-screening** method: it achieves near-perfect sensitivity (TPR ≈ 1) at the cost of selecting more variables than DART VC-measure. It shares the same interface, including `mode` and `backend` arguments.

```{r vip, message = FALSE}
library(bayesVC)

set.seed(123)
n <- 1000
p <- 100
Lrep <- 10

X <- matrix(runif(n * p), n, p)
y_mu <- 10 * sin(pi * X[, 1] * X[, 2]) + 20 * (X[, 3] - 0.5)^2 + 10 * X[, 4] + 5 * X[, 5]
eps <- rnorm(n, mean = 0, sd = 2)
y <- y_mu + eps

VIP_result <- DartVIP(
  y = y,
  X = X,
  seed = 123,
  Lrep = Lrep,
  backend = "BART",
  mode = "regression"
)

VIP_result$pos_idx
```

## Parallel processing

Both `DartVC` and `DartVIP` run the `Lrep` replications via [`future.apply::future_lapply`](https://future.apply.futureverse.org/), so parallelism is controlled by the [`future`](https://future.futureverse.org/) package. By default they run sequentially. To use all available cores, set a `multisession` plan before calling either function:

```{r parallel, message = FALSE, eval = FALSE}
library(future)
library(bayesVC)

# Use all available cores
plan(multisession)

set.seed(123)
n <- 1000
p <- 100
Lrep <- 10

X <- matrix(runif(n * p), n, p)
y_mu <- 10 * sin(pi * X[, 1] * X[, 2]) + 20 * (X[, 3] - 0.5)^2 + 10 * X[, 4] + 5 * X[, 5]
eps <- rnorm(n, mean = 0, sd = 2)
y <- y_mu + eps

VC_result <- DartVC(
  y = y,
  X = X,
  seed = 123,
  Lrep = Lrep,
  backend = "BART",
  mode = "regression"
)
VC_result$pos_idx

# Restore sequential execution
plan(sequential)
```

Note: the `dartMachine` backend does not support `plan(multisession)` because JVM parameters are not inherited by worker processes. Use `plan(sequential)` (the default) when using the `dartMachine` backend.

## Reproduce paper results

### Submit simulations

Experiments for each method can be launched by running the bash script files from the [experiment/](experiment/) folder. For instance, running
```{bash eval = FALSE}
bash submit_DART_VC-measure.sh
```
within the [experiment/](experiment/) folder will submit all simulations for the DART VC-measure method. Raw results, e.g., VC and VC Rank matrices for DART VC-measure, will be stored as JSON files under the [results/](results/) folder.

### Collect raw results

When all raw simulation results are available, simply run the Jupyter notebook [postprocessing/json_collection.ipynb](postprocessing/json_collection.ipynb) to collect all JSON files and prepare them for the variable selection step. Collected results will be stored in [results/no_idx/](results/no_idx/) as `.feather` files.

### Perform variable selection

The Jupyter notebook [postprocessing/variable_selection.ipynb](postprocessing/variable_selection.ipynb) performs variable selection for some of the methods, e.g., the original DART, DART VC-measure, etc., and calculates the selection accuracy for all methods. Selection accuracy results and the selected variables for each simulation and method are stored as `.feather` files under [results/with_idx/](results/with_idx/)

### Reproduce paper figures

R functions in the [postprocessing/](postprocessing/) folders reproduce all figures in the paper. For instance, [postprocessing/fig3_12_13.R](postprocessing/fig3_12_13.R) reproduces Figures 3, 12, and 13 in the paper.


## References

Ye, S. and Li, M. (2025) Posterior Summarization for Variable Selection in Bayesian Tree Ensembles. [arXiv](https://arxiv.org/abs/2509.07121)

Ye, S. and Li, M. (2025) Ab Initio Nonparametric Variable Selection for Scalable Symbolic Regression with Large $p$. *ICML 2025*. [OpenReview](https://openreview.net/forum?id=9gyJJw8ZUj)
